---
title: "Prediction Assignment"
author: "Sylvia Nunez"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
install.packages("ggplot2", dependencies = TRUE)
```


```{r}
library(ggplot2)
library(caret)
library(rpart)
library(rpart.plot)
library(randomForest)
library(corrplot)
```

## Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).


Data

The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv


The data for this project come from this source: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.

## Download and Process Data
```{r}
trainUrl <-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
trainFile <- "./data/pml-training.csv"
testFile  <- "./data/pml-testing.csv"
if (!file.exists("./data")) {
  dir.create("./data")
}
if (!file.exists(trainFile)) {
  download.file(trainUrl, destfile=trainFile, method="curl")
}
if (!file.exists(testFile)) {
  download.file(testUrl, destfile=testFile, method="curl")
}
```



```{r}
trainRaw <- read.csv("./data/pml-training.csv")
testRaw <- read.csv("./data/pml-testing.csv")
dim(trainRaw)
```

```{r}
dim(testRaw)
```

There are 19622 observations and 160 variables in the training data set.
The testing data set has 20 observations and 160 variables. 
We are interested in predicting the "classe" variable in the training set.


Next, we will clean the data.
Get rid of observations with meaningless and missing values.

```{r}
sum(complete.cases(trainRaw))
```

Remove NA's
```{r}
trainRaw <- trainRaw[, colSums(is.na(trainRaw)) == 0]
testRaw <- testRaw[, colSums(is.na(testRaw)) == 0]
```

Remove non-accelorometer measurements.

```{r}
classe <- trainRaw$classe
trainRemove <- grepl("^X|timestamp|window", names(trainRaw))
trainRaw <- trainRaw[, !trainRemove]
trainCleaned <- trainRaw[, sapply(trainRaw, is.numeric)]
trainCleaned$classe <- classe
testRemove <- grepl("^X|timestamp|window", names(testRaw))
testRaw <- testRaw[, !testRemove]
testCleaned <- testRaw[, sapply(testRaw, is.numeric)]
```


We are left with 19622 observations and 53 variables in the clean training data set, 20 observations and 53 variables in the testing data set.


Now we can split the cleaned training set into a pure training data set (70%) and a validation data set (30%). We will use the validation data set to conduct cross validation in future steps.

```{r}
set.seed(23351) # reproducibility
inTrain <- createDataPartition(trainCleaned$classe, p=0.70, list=F)
trainData <- trainCleaned[inTrain, ]
testData <- trainCleaned[-inTrain, ]
```

Data Modeling

```{r}
controlRf <- trainControl(method="cv", 5)
modelRf <- train(classe ~ ., data=trainData, method="rf", trControl=controlRf, ntree=250)
modelRf
```
Seen above, using Random Forest, we fit a predictive model for activity recognition algorithm, we will use 5-fold cross validation when applying the algorithm.


Now, we estimate the performance of the model on the validation data set.
```{r}
predictRf <- predict(modelRf, testData)
confusionMatrix(as.factor(testData$classe), predictRf)
```

```{r}
accuracy <- postResample(predictRf, as.factor(testData$classe))

accuracy
```

```{r}
outofsampleerror <- 1 - as.numeric(confusionMatrix(as.factor(testData$classe), predictRf)$overall[1])

outofsampleerror
```
We observe,the estimated accuracy of the model is 99.40% and the estimated out-of-sample error is 0.58%.


### Predicting for Test Data Set
Now, we apply the model to the original testing data set downloaded from the data source. We remove the problem_id column first.

```{r}
result <- predict(modelRf, testCleaned[, -length(names(testCleaned))])
result
```

